---
title: "PeerAssignment1"
author: "Venu Palvai"
date: "Sunday, July 19, 2015"
output: html_document
---

This document shows the steps for executing PeerAssignment1 to analyze personal
activity data of a sample individual.

[Download unzip file from this location] [1]

[1]: https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip

##Loading and Pre-Processing data

``` {r, echo=TRUE}

activity_data <- read.csv("activity.csv" 
                          ,colClasses = c("integer", "Date", "factor"))
activity_data$month <- as.numeric(format(activity_data$date, "%m"))
activity_data_noNA <- na.omit(activity_data)
rownames(activity_data_noNA) <- 1:nrow(activity_data_noNA)
head(activity_data_noNA)
dim(activity_data_noNA)
library(ggplot2)

```
## What is mean total number of steps taken per day?
``` {r, echo=TRUE}

steps_by_day <- aggregate(steps ~ date, activity_data_noNA, sum)
qplot(steps_by_day$steps, binwidth=1000, xlab="Total number of steps taken each day")

rmean <- mean(steps_by_day$steps)
rmedian <- median(steps_by_day$steps)
```
Mean is `r rmean`  
Median is `r rmedian`

## What is the average daily activity pattern?

``` {r, echo=TRUE}
steps_by_interval <- aggregate(steps ~ interval, activity_data_noNA, mean)
ggplot(data=steps_by_interval, aes(x=interval, y=steps)) +
    geom_line() +
    xlab("5-minute interval") +
    ylab("average number of steps taken")


```
## Imputing missing values
``` {r, echo=TRUE}
total_NA <- sum(is.na(activity_data))
```
Total rows with NA values `r sum(is.na(activity_data))`

Create a new dataset with missing values filled in 
loop the original dataset
check for missing values, fill in the values with avgsteps for the interval
``` {r, echo=TRUE}
new_activity_data <- activity_data 
for (i in 1:nrow(new_activity_data)) {
    if (is.na(new_activity_data$steps[i])) {
        new_activity_data$steps[i] <- steps_by_interval[which(new_activity_data$interval[i] == steps_by_interval$interval), ]$steps
    }
}

head(new_activity_data)
sum(is.na(new_activity_data))
```
Plotting new data with missing values replaced by avg interval steps
``` {r, echo=TRUE}
new_steps_by_day <- aggregate(steps ~ date, new_activity_data, sum)
qplot(new_steps_by_day$steps, binwidth=1000, xlab="Total number of steps taken each day (No NA data)")

new_rmean <- mean(new_steps_by_day$steps)
new_rmedian <- median(new_steps_by_day$steps)


```
New Mean is `r new_rmean`  
New Median is `r new_rmedian`

## Are there differences in activity patterns between weekdays and weekends?
``` {r, echo=TRUE}
new_activity_data$weekdays <- factor(format(new_activity_data$date, "%A"))
levels(new_activity_data$weekdays)
levels(new_activity_data$weekdays) <- list(weekday = c("Monday", "Tuesday",
                                             "Wednesday", 
                                             "Thursday", "Friday"),
                                 weekend = c("Saturday", "Sunday"))
levels(new_activity_data$weekdays)
table(new_activity_data$weekdays)

new_averages <- aggregate(steps ~ interval + weekdays, data=new_activity_data, mean)
library(lattice)
xyplot(new_averages$steps ~ new_averages$interval | new_averages$weekdays, 
       layout = c(1, 2), type = "l", 
       xlab = "Interval", ylab = "Number of steps")
```
